Designing Types for R, Empirically

### 2020-05-15 paper/main.tex
merge

\AT{Me}

In the last section, we evaluated our type annotation language based on a set of function types that were automatically generated by \typetracer.
Now, we're interested in knowing how representative our trace-generated types are of the functions they were generated for.
To measure this, we conducted a large-scale experiment: for each package in the corpus discussed in Section~\ref{sec:corpus}, we ran \typetracer on its test, example, and vignette code, obtaining types for package functions, and then subsequently ran the test, example, and vignette code of each of the package's reverse dependencies (or clients).
We collected failed contract assertions in order to analyze them and perhaps identify weaknesses in our type annotation language.

@@

We ran our evaluation on \NUMPKGSEVAL packages and recorded \NUMASSERTIONS total assertions.
Overall, we found that only \PERCFAILEDASSERTIONSWUNDEF of contract assertions failed.
We made a simplification in our analysis, capping the number of function arguments we generate types for to 20, and this resulted in only \PERCASSERTIONSUNDEF failed assertions, which suggests that the simplification we made did not greatly affect the results.
Once we controlled for these, we found that \PERCFAILEDASSERTIONS of all assertions failed for reasons related to our generated types.
Additionally, we found that \PERCSUCCARG of parameter types and \PERCSUCCFUNS of functions types were never violated.
The number of immaculate function types increases to \PERCSUCCFUNNOSTHREE if we count out S3 functions:
These S3 functions represent user-defined polymorphism which we do not tackle in this work.
Overall, these numbers are promising, and suggest that the type signatures generated by running \typetracer on our corpus are indeed representative, and capture intended function behaviour.

@@

We breakdown the failed assertions by type in Table~\ref{tbl:fail-breakdown-0}.
By far the most commonly failed assertion is one asserting that a value with ${\bf dbl}[]$ type has type ${\bf class}\langle matrix \rangle$.
This was a deliberate design decision, as coercion from vectors to matrices is ad hoc at best, and surely not a practice codified in the language.
In a similar vein, another popular failing assertion is checking if a ${\bf dbl}[]$ has type ${\bf int}[]$, another case of commonly performed coercion.
We did not include these types of coercions in our type annotation framework as programmers cannot rely on them, and it is not always the case that the coercions are safe to perform.

@@

In addition to the raw number of failing contract checks, we were interested in how many functions had a parameter where a contract check failed.
Overall, we found that \PROPFUNSFAILEDCHECK of functions had a failing contract check.
Digging a little deeper, we noticed that many of these functions were performing S3 dispatch, which is part of R's object-orientation framework which we will tackle in future work.
Removing those functions, we see that \PROPFUNSFAILEDCHECKNOSTHREE of functions exhibited failing contract checks.
These are functions which were under-tested, as calls to these functions represent only \PERCCALLSBADFUNSINTESTS of recorded calls during \typetracer's run on the core corpus to infer types.

>>>>

In the last section, we mainly discussed the function types that \typetracer generated, focusing on analyzing the polymorphism present in the types.
Now, we're interested in knowing how representative our trace-generated types are of the functions they were generated for.
To measure this, we conducted a large-scale experiment: for each package in the corpus discussed in Section~\ref{sec:corpus}, we ran \typetracer on its test, example, and vignette code, obtaining types for package functions, and then subsequently ran the test, example, and vignette code of each of the package's reverse dependencies.
We collected failed contract assertions in order to analyze them and perhaps identify weaknesses in our type annotation language.

@@

We ran our evaluation on \NUMPKGSEVAL packages and recorded \NUMASSERTIONS total assertions.
Overall, we found that only \PERCFAILEDASSERTIONSWUNDEF of contract assertions failed.
We made a simplification in our analysis, capping the number of function arguments we generate types for to 20, and this resulted in only \PERCASSERTIONSUNDEF of assertions failing: this suggests that the simplification we made did not greatly affect the results.
Once we controlled for these, we found that \PERCFAILEDASSERTIONS of all assertions failed for reasons related to our generated types.
Additionally, we found that \PERCSUCCARG of parameter types and \PERCSUCCFUNS of functions types were never violated.
The number of immaculate function types increases to \PERCSUCCFUNNOSTHREE if we count out S3 functions:
These S3 functions represent user-defined polymorphism which we do not tackle in this work.
Overall, these numbers are promising, and suggest that the type signatures generated by running \typetracer on our corpus are indeed representative, and capture intended function behaviour.

@@

We breakdown the failed assertions by type in Table~\ref{tbl:fail-breakdown-0}.
By far the most commonly failed assertion is one asserting that a value with \VEC\dbl type has type \CLASS{matrix}.
This was a deliberate design decision, as coercion from vectors to matrices is ad hoc at best, and surely not a practice codified in the language.
In a similar vein, another popular failing assertion is checking if a \VEC\dbl has type \VEC\int, another case of commonly performed coercion.
We did not include these types of coercions in our type annotation framework as programmers cannot rely on them, and it is not always the case that the coercions are safe to perform.

@@

In addition to the raw number of failing contract checks, we were interested in how many functions had a parameter where a contract check failed, and overall we found this to be the case in \PROPFUNSFAILEDCHECK of functions.
Digging a little deeper, we noticed that many of these functions were performing S3 dispatch, which is part of R's object-orientation framework: thus, these functions implement user-defined polymorphism which we do not handle in our type framework.
Removing those functions, we see that \PROPFUNSFAILEDCHECKNOSTHREE of functions exhibited failing contract checks.
These are functions which were under-tested, as calls to these functions represent only \PERCCALLSBADFUNSINTESTS of recorded calls during \typetracer's run on the core corpus to infer types.
